---
title: Predicting Property Value within the Liverpool Authority District using Spatial Econometric Models

# Use letters for affiliations
author:
  - name: 201374125
    affiliation: a
address:
  - code: a
    address: Department of Geography and Planning, University of Liverpool, Liverpool, L69 7ZX

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: 201374125

# Place DOI URL or CRAN Package URL here
#doi: "https://cran.r-project.org/package=YourPackage"

# Abstract
abstract: |
 Word Count: 1643

# Paper size for the document, values of letterpaper and a4paper
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography
bibliography: /home/cjber/Dropbox/bib/zlib.bib
biblio-style: apsr

# Optional: Enable a 'Draft' watermark on the document
watermark: false

# Customize footer, eg by referencing the vignette
footer_contents: "201374125"

# Produce a pinp document
output: pinp::pinp

# Required: Vignette metadata for inclusion in a package.
vignette: >
  %\VignetteIndexEntry{YourPackage-vignetteentry}
  %\VignetteKeywords{YourPackage, r, anotherkeyword}
  %\VignettePackage{YourPackage}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  cache = TRUE,
  echo = FALSE,
  results = FALSE,
  warning = FALSE
)
```
```{r, libs}
library(tidyverse)
library(ggplot2)
library(sf)
library(tmap)
library(tmaptools)
library(ggmap)
library(smoothr)
library(extrafont)
library(ggspatial)
library(gstat)
library(GISTools)
library(raster)
library(rasterVis)
library(broom)
library(kableExtra)
library(viridis) # colour schemes
library(spdep)
```
```{r, shps}
property <- read_sf("./data/house_transactions/liv_house_trans.shp")
property$price_log <- log(property$price)
property <- property[property$type != "O", ] # exclude 'other', removes very expensive non residential sales

liv <- read_sf("./data/house_transactions/liv_outline.shp")
imd <- read.csv("./data/house_transactions/E08000012.csv")

lsoa <- read_sf("./data/Liverpool_lsoa11.shp")
```
```{r, theme}
theme_map <- function(...) {
  theme_minimal() +
    theme(
      axis.line = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.border = element_rect(colour = "grey50", fill = NA, size = 0.5),
      ...
    )
}
```

# Introduction

The importance of the spatial element in determining the factors driving property price is widely recognised [@bitter2007], and controlling for this in the production of property pricing models is essential. The spatial variation in property price is driven by two key econometric issues;

1. Spatial Dependence

> The lack of independence often present in observations of cross-sectional data, i.e. the property price is influenced by surrounding property prices

2. Spatial Heterogeneity

> The underlying process cannot be assumed to be constant throughout the geography selected, i.e. the variation in property prices relies on an aggregate of geographical features that vary across space [@anselin2013;@anselin2013a]

Extending these concepts in relation to property prices will allow for an improved understanding of these key spatial processes that are not always considered in spatial models in property price research [@bowen2001]. Typically empirical models of property price research have treated the chosen area (often metropolitan) as a single market [@bitter2007], although recent developments in Multi Level models for use in spatial analysis have extended the analysis, allowing for the introduction of the concept of housing 'neighbourhoods' [e.g. @dong2019].

This report will first utilise a spatial fixed effect model, which will separate each individual area as a fixed effect (in this case two digit postcodes). The fixed effect model therefore will compare only properties that share the same postcode, isolating the effects of the chosen predictor variables. This model essentially allows the intercept to vary between the two digit postcodes within Liverpool [@anselin2013a]. Secondly a Spatial Lag model will be used to assess the degree of spatial dependence that is observed in a selected variable.

```{r, kde}
## Function to choose bandwidth according to Bowman and Azalini / Scott's rule
## Adapted for sf
choose_bw <- function(spdf) {
  X <- st_coordinates(spdf)
  sigma <- c(sd(X[, 1]), sd(X[, 2])) * (2 / (3 * nrow(X)))^(1 / 6)
  return(sigma / 1000)
}

kde_property <- smooth_map(property, cover = liv, bandwidth = choose_bw(property))

area_threshold <- units::set_units(1000, km^2)
kde_property$polygons <- kde_property$polygons %>%
  st_transform(4326) %>%
  fill_holes(area_threshold)

bg <- liv %>%
  st_transform(4326) %>%
  st_union() %>%
  fill_holes(threshold = area_threshold)

# change level names to colours for slight ease
bins <- kde_property$polygons$level
bincol <- c("#ebebe5", "#dfe3ee", "#8b9dc3", "#3b5998")
kde_property$polygons$level <- bincol

kde_map <- ggplot() +
  annotation_map_tile(data = bg) +
  geom_sf(data = bg, size = 1, col = "black", fill = NA) +
  geom_sf(
    data = kde_property$polygons,
    aes(col = bins, fill = bins),
    color = "black", alpha = 0.75
  ) +
  scale_fill_manual(values = bincol) +
  theme_map() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.spacing.x = unit(.1, "cm"),
    legend.text = element_text(margin = margin(t = 0), size = 5)
  ) +
  guides(fill = guide_legend(
    title = "Cyl",
    label.position = "left"
  ))
```

# Preliminary Analysis

Prior to running the models, a Kernel Density Map was produced which allowed for an exploration of the distribution of property sales (Figure \ref{fig:kde_fig}). To select the bandwidth an adapted version of Scott's rule was implemented into $r$ [@bowman1997]:

```{r, eval = FALSE, echo = TRUE}
choose_bw <- function(spdf) {
  X <- st_coordinates(spdf)
  sigma <- c(sd(X[, 1]), sd(X[, 2])) *
    (2 / (3 * nrow(X)))^(1 / 6)
  return(sigma / 1000)
}
```

```{r, kde_fig, fig.cap="Kernel Density Estimate showing property sales in the Liverpool region. Bandwidth selected using Scott's Rule"}
kde_map
```
Figure \ref{fig:kde_fig} shows that a high proportion of property transactions occur around the city centre, with fewer on the outskirts. Likely due to there being a higher density of properties in the city centre, and more flats as opposed to houses.

```{r, idw}
property_sp <- property %>%
  st_transform(4326) %>%
  as_Spatial()

bg_sp <- bg %>%
  as_Spatial()

set.seed(9)
# Create an empty grid where n is the total number of cells
grd <- as.data.frame(spsample(bg_sp, "regular", n = 50000))
names(grd) <- c("X", "Y")
coordinates(grd) <- c("X", "Y")
gridded(grd) <- TRUE # Create SpatialPixel object
fullgrid(grd) <- TRUE # Create SpatialGrid object

# Add P's projection information to the empty grid
proj4string(grd) <- proj4string(property_sp)

# Interpolate the grid cells using a power value of 2 (idp=0.8)
P.idw <- idw(price_log ~ 1, property_sp, newdata = grd, idp = 0.8)

# convert to raster, clip to liverpool
r <- raster(P.idw)
r.m <- mask(r, bg_sp)
```
```{r, idw_fig}
# need to fortify to plot tile and border on same ggplot
bg_points <- fortify(bg_sp)
idw_map <- gplot(r.m) +
  annotation_map_tile(data = bg) +
  geom_raster(aes(fill = value), interpolate = TRUE, alpha = .8) +
  geom_path(data = bg_points, aes(x = long, y = lat), size = 1) +
  scale_fill_gradientn(
    colours = c("white", "red", "black"), limits = c(11.24, 11.85),
    name = " ", na.value = NA
  ) +
  theme_map() +
  theme(
    legend.key.width = unit(1, "cm"),
    legend.position = "bottom",
    legend.spacing.x = unit(.1, "cm"),
    legend.text = element_text(margin = margin(t = 0.5), size = 5),
  )
```

```{r, fig_idw, fig.cap="Spatial Interpolation of Property Sales Price within Liverpool using Inverse Distance Weighting"}
idw_map
```

Figure \ref{fig:fig_idw} gives the results of a spatial interpolation technique showing the estimated distribution of property value throughout the LAD. North of the city centre there tends to be lower property values, whereas south of the city centre there is a band of high property values.

# Regression Models

```{r}
property$pc <- as.character(lapply(
  strsplit(as.character(property$pcds), split = " "), "[", 1
))

property <- merge(property, imd)
```

```{r}
## convert p vals to asterisks
p_val <- function(x) {
  symnum(x, corr = FALSE, na = FALSE, cutpoints = c(
    0, 0.001, 0.01, 0.05, 1
  ), symbols = c("***", "**", "*", " "))
}
```

## Baseline non-spatial regression

The results from the preliminary OLS regression analysis given on Table \ref{tab:ols_table} show that each chosen predictor variable significantly influences the property price. Particularly, flats negatively influence the property price by the largest factor; flats are associated with a 0.84% decrease in the value of property price. Similarly Semi-detached and terraced properties are associated in a price reduction, suggesting that detached properties are associated with a significant price increase.

Notably the Index of Multiple Deprivation (IMD) Score does negatively influence property price significantly, but only to a small degree, likely as IMD Score is an aggregate of many deprivation indicators, and the other chosen variables may be acting as a proxy towards these. Additionally it should be noted that IMD is an area level variable, and associated with the Lower Super Output Area (LSOA) where the property is located, therefore the assumption is made that every property is associated with the IMD Score of the containing LSOA. The equation for the OLS regression is given below, (Equation \ref{eq:ols}), house prices were log transformed as without log transformation the distribution was not normal, an assumption with OLS regression [@li2012].

\begin{equation}\label{eq:ols}
\log{P_i} = \alpha + \beta_1 NEW_i + \beta_2 IMD_i + \beta_3 TYPE_i + \epsilon_i
\end{equation}


```{r, ols, echo=F, results=T}
m1 <- lm("price_log ~ new + imd_score + type", property)

# tidy from broom package changes summary information into data.frames
ols_m1 <- tidy(summary(m1))
# select only the important columns

# Change p values into vectors for each model
p <- p_val(ols_m1$p.value)

# Round statistic values to 2 digits and change to vector as well, keep zeros
t <- sprintf("%.2f", ols_m1$estimate)

# Join statistic values and p asterisks
ols_m1$estimate <- paste(t, p)

# Find R squared % for each model
r <- summary(m1)$r.squared * 100

# Round them to 2 digits
r <- round(r, 2)
# rename row names of table
ols_m1$term <- c("(Intercept)", "New Build", "IMD Score", "Flat", "Semi-Detached", "Terraced")

ols_m1 <- ols_m1[1:3]
```

```{r, ols_table, echo=F, results=T}
# tell Kable to not show anything for NA values
options(knitr.kable.NA = "")
kable(ols_m1,
  digits = 2, caption = "OLS Linear Regression Model",
  linesep = "",
  longtable = FALSE, booktabs = TRUE,
  format = "latex",
  align = c("l", "c", "c", "c"),
  col.names = c(
    "Variable",
    "Model Estimate",
    "Standard Errors"
  )
) %>%
  # footnote for significance
  footnote(
    general_title = "",
    general = c(
      paste("† R Squared =", r),
      "*** Significant at the 0.001 level"
    )
  )
```

## Spatial Fixed Effects Model based on 2 Digit Postcode

OLS regression assumes that all properties share the same constant term and alpha across all spatial areas within Liverpool, however, space can be included into the model, to act as a proxy for unobserved variables that act vary across space. The equation for this model is given (Equation \ref{eq:fe}). In this model, property 'type' is converted into the dummy variable 'detached' which identifies whether a particular property is detached and given the value 1, otherwise given 0. Similarly the category 'new' has been changed to a value of either 1 or 0.

\begin{equation}\label{eq:fe}
\log{P_i} = \alpha_r + \beta_1 NEW_i + \beta_2 IMD_i + \beta_3 DETACHED_i + \epsilon_i
\end{equation}

```{r}
property$one <- 1
property$detached <- 0
property[property$type == "D", "detached"] <- 1
property$newB <- 1
property[property$new == "N", "newB"] <- 0

m2 <- lm("price_log ~ pc + newB + detached + imd_score - 1", property)

# tidy from broom package changes summary information into data.frames
t_m2 <- tidy(summary(m2))
# select only the important columns
tail(t_m2)

# Change p values into vectors for each model
p <- p_val(t_m2$p.value)

# Round statistic values to 2 digits and change to vector as well, keep zeros
t <- sprintf("%.2f", t_m2$estimate)

# Join statistic values and p asterisks
t_m2$estimate <- paste(t, p)

# Find R squared % for each model
r <- summary(m2)$r.squared * 100

# Round them to 2 digits
r <- round(r, 2)
# rename row names of table


t_m2$term <- as.character(lapply(
  strsplit(as.character(t_m2$term), split = "pc"), "[", 2
))
t_m2 <- t_m2[1:3]

val <- c("New", "Detached", "IMD Score")

t_m2$term[is.na(t_m2$term)] <- val
```

```{r, fe_table, echo=F, results=T}
# tell Kable to not show anything for NA values
options(knitr.kable.NA = "")
kable(t_m2,
  digits = 2, caption = "Spatial Fixed Effects Model",
  linesep = "",
  longtable = FALSE, booktabs = TRUE,
  format = "latex",
  align = c("l", "c", "c", "c"),
  col.names = c(
    "Variable",
    "Model Estimate",
    "Standard Errors"
  )
) %>%
  # footnote for significance
  footnote(
    general_title = "",
    general = c(
      paste("† R Squared =", r),
      "*** Significant at the 0.001 level"
    )
  )
```

Here space is included into the model as it is known to have a significant influence on property price [@waddell1993]. Notably the accessibility of an area in terms of general connectivity with the city is known to drive an increase in property value [@narvaez2012]. As well as the access to amenities in the immediate area, including public transport connections [@bartholomew2011]. In this model therefore, the inclusion of two digit postcodes will act as 'neighbourhoods' in which the baseline property value is expected to vary due to the variables noted above, in addition to spatial characters that are difficult to quantify. Essentially the spatial heterogeneity of Liverpool is now considered. Table \ref{tab:fe_table} indicates that the postcode L18 is the most desirable postcode as it displays the highest log property price intercept ($\text{Model Estimate} = 12.31$) when detached homes, new builds, and IMD score is taken into account. Figure \ref{fig:map_int} displays these intercepts graphically, which gives an indication as to where the highest and lower intercepts are located. The high $r^2$ value shown on table \ref{tab:fe_table} likely reflects only the high number of variables, and doesn't assist with model fit interpretation. It is clear from Figure \ref{fig:map_int} that property price is spatially correlated, and as such likely is effected by spatial dependence.

```{r}
int_map <- tidy(summary(m2))

int_map$pc <- as.character(lapply(
  strsplit(as.character(int_map$term), split = "pc"), "[", 2
))

int_map <- merge(property, int_map, by = "pc")
int_map <- lsoa %>%
  st_join(int_map) %>%
  group_by(LSOA11CD.x) %>%
  summarise(estimate = mean(estimate))
```
```{r, quantiles}
no_classes <- 6 # specify 6 bins
labels <- c() # empty vector

# find quantiles for each msoa
quantiles <- quantile(int_map$estimate,
  probs = seq(0, 1, length.out = no_classes + 1), na.rm = T
)

# define custom labels, remove decimal etc
labels <- c()
for (idx in 1:length(quantiles)) {
  labels <- c(labels, paste0(round(quantiles[idx + 1], 2)))
}

# remove final label
labels <- labels[1:length(labels) - 1]

# new variable in dataset with quantiles
int_map$estimate_qu <- cut(int_map$estimate,
  breaks = quantiles,
  labels = labels,
  include.lowest = T
)
```

```{r}

int_map <- st_transform(int_map, 4326)
# remove slivers
int_map <- st_buffer(int_map, dist = 0.001)

map_int <- ggplot() +
  annotation_map_tile(data = bg) +
  geom_sf(data = int_map, aes(fill = estimate), colour = NA, size = 0) +
  geom_path(data = bg_points, aes(x = long, y = lat), size = 1) +
  theme_map() +
  theme(
    legend.key.width = unit(1, "cm"),
    legend.position = "bottom",
    legend.spacing.x = unit(.1, "cm"),
    legend.text = element_text(margin = margin(t = 0.5), size = 5)
  ) +
  scale_fill_viridis(
    option = "magma",
    direction = -1,
    name = " "
  )
```
```{r, map_int, fig.cap="Varying Spatial Estimates for Log property Price, produced by the Spatial Fixed Effects Model"}
map_int
```

## Including Spatial Lag of IMD Score

The inclusion of the spatial lag of IMD Score into the model builds on the assumption that property value within a certain postcode is influenced by the IMD Score of surrounding properties (spatial dependence). It should be noted again that IMD Score is an LSOA level unit, so many properties share the exact same IMD Score, for this reason, the property sale prices were aggregated and averaged into each LSOA, queen weightings were used to demonstrate the lag in IMD Score between neighbouring LSOAs. Again, Table \ref{tab:lag} reveals that detached properties have by far the greatest positive influence on property prices (Detached properties indicate a 0.64% rise in property price). Notably the spatial lag for IMD score is significant, but gives a relatively small negative value, potentially indicating either the lack of influence that IMD Spatial lag has, or that the new and detached properties dummy variables are acting as a proxies towards the majority of IMD Score indicators.

\begin{equation}\label{eq:lag}
y_{lag-i} = \displaystyle \sum_j w_{ij} y_j
\end{equation}

\begin{equation}
\log{P_i} = \alpha + \beta_{1} NEW_i + \beta_{2} DETACHED_i + \beta_{3} IMD_i + \rho IMD_{lag-i} + \epsilon_i
\end{equation}

```{r}
# aggregate by lsoa for imd score inclusion
prop_lsoa <- lsoa %>%
  st_join(property, by = LSOA11CD) %>%
  group_by(LSOA11CD.x) %>%
  summarise(
    price_log = mean(price_log),
    newB = sum(newB) / length(newB),
    imd_score = mean(imd_score),
    detached = sum(detached / length(detached))
  ) %>%
  na.omit()

# create queen weights for imd score using lsoa
lsoa_w <- poly2nb(prop_lsoa)
lsoa_lw <- nb2listw(lsoa_w)
prop_lsoa$imd_score_w <- lag.listw(lsoa_lw, prop_lsoa$imd_score)
```

```{r}
m4 <- lm("price_log ~ newB + detached + imd_score + imd_score_w", prop_lsoa)
```
```{r}
# spatial lag of imd_score

property$imd_score_w <- lag.listw(hknn, property$imd_score)
```

```{r}
m4 <- lm("price_log ~ new + detached + imd_score + imd_score_w", property)
```

```{r}
# tidy from broom package changes summary information into data.frames
sp_m4 <- tidy(summary(m4))
# select only the important columns

# Change p values into vectors for each model
p <- p_val(sp_m4$p.value)

# Round statistic values to 2 digits and change to vector as well, keep zeros
t <- sprintf("%.2f", sp_m4$estimate)

# Join statistic values and p asterisks
sp_m4$estimate <- paste(t, p)

# Find R squared % for each model
r <- summary(m1)$r.squared * 100

# Round them to 2 digits
r <- round(r, 2)

# rename row names of table
sp_m4$term <- c("(Intercept)", "New Build", "Detached", "IMD Score", "IMD Score (Lag)")

sp_m4 <- sp_m4[1:3]
```

```{r, lag, echo=F, results=T}
# tell Kable to not show anything for NA values
options(knitr.kable.NA = "")

kable(sp_m4,
  digits = 2, caption = "Spatial Lag Model",
  linesep = "",
  longtable = FALSE, booktabs = TRUE,
  format = "latex",
  align = c("l", "c", "c", "c"),
  col.names = c(
    "Variable",
    "Model Estimate",
    "Standard Errors"
  )
) %>%
  # footnote for significance
  footnote(
    general_title = "",
    general = c(
      paste("† R Squared =", r),
      "*** Significant at the 0.001 level"
    )
  )
```

# Discussion

Spatial interpolation is able to alleviate the problem scale and resolution has in relation to the modifiable aerial unit problem [MAUP; @openshaw1984] if points were instead aggregated into spatial units such as LSOA [@fotheringham1989]. LSOA units are broadly unrelated to the outcome variable property price, whereas interpolation uses the underlying data to estimate the distribution. While spatial interpolation tackles the issues with the MAUP, it introduces more assumptions, namely the assumption that property prices are all spatially correlated with each other, to the same degree, and that the distance decay selected with the calculation appropriately reflects that of the underlying, but missing, data [@kar2012].

The figure displaying the spatial interpolation of property prices (Figure \ref{fig:fig_idw}) displays both spatial heterogeneity, where high and low values are clustered spatially, both towards the north and south of Liverpool, and spatial dependence as property price exhibits gradual shifts from high to lower values [See also \ref{fig:fig_idw};@anselin2013]. The presence of these two econometric issues means that without considering space, predicting house price would be inaccurate purely by using variables in a non spatial model as they violate the underlying assumptions of regression analysis [@anselin2010].

Spatial fixed effects allow the properties within postcodes to be treated independently, attempting to remove the spatial correlations between postcodes to achieve a more accurate prediction of property price. However, it is limited in that once the number of spatial groups (e.g. Postcodes) become too large, the model itself becomes too large to appropriately interpret. Additionally, the $r^2$ value for the spatial fixed effect model chosen adds little information in terms of the model fit, essentially the model fit is often hard to interpret and quantify, so must be judged independently. While it is often suggested that spatial fixed effects models do remove spatial correlations, @anselin2013a note that with true spatial dependence this may not be true. As spatial fixed effects use space as a proxy for many unobserved variables, it is difficult to determine exactly what causes the between group differences, and due to this, any attempt to utilise the model outside of the specified area will not work.

Alternative methods for treating spatial dependence and heterogeneity exist, for example @mcmillen2010 suggests that any observed spatial autocorrelation in models is due to omitting variables, and states that non spatial semi-parametric modelling should be used instead of using spatial fixed effects as a proxy for unobserved variables.

The primary issue with the inclusion of the IMD Score with this analysis is that IMD Score is automatically aggregated into LSOAs within Liverpool, and so is an area level indicator paired with individual level indicators such as property type, new builds, and price, and an assumption of spatial fixed effects is that all variables chosen are at an individual level [@anselin2013].

A multi level model approach to determining the area effect of house prices could consider the postcodes as a lower aggregation unit, and the entire LAD as the higher level unit. Results determined from a multi level analysis would indicate the level of variation between the two units, allowing for the consideration that housing spatial and temporal trends occur at more than just a macro level [@gibbons2008]. Additionally, the majority of multi level models in property price research find that there is at least some unexplained variation between lower and higher level hierarchical units [@dong2019; @jones1993], suggesting that there is an area effect that determines spatial differences in property price. The models utilised above however give insight into the spatial dependence and heterogeneity that is apparent with property prices within Liverpool.
